---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# sostools

<!-- badges: start -->
<!-- badges: end -->

This is just a collection of functions I use in my research that I use
frequently. Other functions I use infrequently can be found in my `snippets`
repo.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tsostarics/sostools")
```
## Easily reference coefficients in prose

Here is a simple example of referencing the values of a model summary
without needing to hand write or format it yourself. See `?get_coef` for more
information. `make_coef_getter` is a function factory that provides a shorthand
if you have multiple models you're working with.

```{r example}
library(sostools)
library(tidyverse)
mdl <- lm(mpg ~ cyl * gear, data = mtcars) # Create model
coefs <- enlist_coefs(mdl) # Create coefficients
get_coef1 <- make_coef_getter(mdl) # Using the model object directly
get_coef2 <- make_coef_getter(coefs) # Using the list of coefficients we made
get_coef("cyl:gear", coefs)
```

Here we can get the formatted values of a write up easily with in-line R code.
We found no effect of cylinder `r get_coef2("cyl")` or gear `r get_coef2("gear")`
nor an interaction `r get_coef2("cyl:gear")`.

## Retain contrast coding labels

```{r}
mdl_data <- 
  mtcars %>% 
  as_tibble() %>% 
  mutate(cyl = factor(cyl), 
         twolevel = factor(round(runif(n()),0)))
contrasts(mdl_data$cyl) # 3 level treatment coding
contrasts(mdl_data$twolevel) # 2 level treatment coding
contrast_code(mdl_data$twolevel) # 2 level scaled sum coding

# note the matrix orientation for 3 level scaled sum coding
contrasts(mdl_data$cyl) <- contrast_code(mdl_data$cyl,
                                        matrix(c(-1/3, 2/3, -1/3,
                                                 -1/3, -1/3, 2/3),
                                               nrow = 3))

contrasts(mdl_data$cyl) # contrasts now updated with labels retained
```

