---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# sostools

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/tsostarics/sostools/branch/master/graph/badge.svg)](https://codecov.io/gh/tsostarics/sostools?branch=master)
<!-- badges: end -->

This is just a collection of functions I use in my research that I use
frequently. Other functions I use infrequently can be found in my `snippets`
repo.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tsostarics/sostools")
```
## Easily reference coefficients in prose

Here is a simple example of referencing the values of a model summary
without needing to hand write or format it yourself. See `?get_coef` for more
information. `make_coef_getter` is a function factory that provides a shorthand
if you have multiple models you're working with.

```{r example}
library(sostools)
library(tidyverse)
mdl <- lm(mpg ~ cyl * gear, data = mtcars) # Create model
coefs <- enlist_coefs(mdl) # Create coefficients
get_coef1 <- make_coef_getter(mdl) # Using the model object directly
get_coef2 <- make_coef_getter(coefs) # Using the list of coefficients we made
get_coef("cyl:gear", coefs)
get_coef1("cyl:gear")
get_coef2("cyl:gear")
```

Here we can get the formatted values of a write up easily with in-line R code.
You could say something like: We found no significant effect of cylinder 
`r get_coef2("cyl")` or gear `r get_coef2("gear")` nor a significant interaction
between the two `r get_coef2("cyl:gear")`.

You can also use these with Bayesian models fit with `brms`, which will give you
the $\beta$ estimate as well as the credible interval.

## Contrast coding functionality

See the `contrasts` vignette with `vignette('contrasts', 'sostools')`.
