---
title: "contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes some of the contrast coding helpers this package provides.
It should be noted that this package does not seek to provide an interface for
creating hypothesis matrices and their respective contrast matrices. For that
functionality you might look into the `multcomp` or `hypr` packages. The contrast
functionality of this package is focused on the following:

- Easily implement commonly used contrast coding schemes
- Make reference levels explicit
- Cut down on repetitive coding involved in setting contrasts
- Retain interpretable labels in model summary coefficients
- Allow the implementation of custom contrasts (but again not their creation)

This vignette will not describe how different contrast schemes are created or
interpreted, but will describe some of the high level differences. I recommend
reading the following:

- [UCLA R Library Contrast Coding Systems for Categorical Variables](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/)
- [How to capitalize on a priori contrasts in linear (mixed) models: A tutorial](https://www.sciencedirect.com/science/article/pii/S0749596X19300695)

```{r setup, message=FALSE, warning=FALSE}
library(sostools)
library(dplyr)
library(MASS)
mdl_data <- 
  mtcars %>% 
  as_tibble() %>% 
  mutate(cyl = factor(cyl), 
         twolevel = round(runif(n()),0),
         twolevel = ifelse(twolevel == 1, "a", "b"),
         twolevel = factor(twolevel),
         gear = factor(gear),
         carb = factor(carb))
```

## Motivation 

The default contrast scheme for factors in R is **treatment coding**, which tests
the difference between each level of a factor and the level set as the reference 
level. Often this doesn't correspond to hypotheses we want to test, so we need to 
change it ourselves.

```{r}
contrasts(mdl_data$twolevel)
contrasts(mdl_data$carb)
```

Typically we would change these manually using `contrasts<-`, for example
this would implement R's built in sum coding for a 6 level factor.

```{r}
contrasts(mdl_data$carb) <- contr.sum(6)
contrasts(mdl_data$carb)
```

Compare this contrast matrix to the treatment coded matrix previously shown
for this factor. There are two things to note here. First, in the treatment
coded matrix the *first row* corresponding to level 1 was all 0, denoting 1 as
the reference level, but here the reference level is actually level 8, which has
all -1s. This may not be what you intended.
Second, the **column names have changed**. The column names represent the labels
for the comparisons we're making with our contrast matrix. In the treatment
coded matrix, the labels are `c(2, 3, 4, 6, 8)`, which would be shown in the
coefficient names in a model summary. In the matrix from `contr.sum(6)`, they're
not the same. Notably, what would correspond to the comparison for level 6 would
be displayed as `carb5` in our model output! We'd like to have meaningful labels
that make it easier to interpret.

Occasionally people will mention that there's a "standard", "tradition", or
"convention" that `coef1` corresponds to something. This isn't really the case,
it's only a "convention" because people are unknowingly erasing or not setting
the labels themselves. There's no reason to need to keep track of what 1 represents
for all your categorical variables in your head, as you'll likely make a mistake.

```{r reset, echo = FALSE}
mdl_data <- 
  mtcars %>% 
  as_tibble() %>% 
  mutate(cyl = factor(cyl), 
         twolevel = round(runif(n()),0),
         twolevel = ifelse(twolevel == 1, "a", "b"),
         twolevel = factor(twolevel),
         gear = factor(gear),
         carb = factor(carb))
```

## Contrast coding 

This package provides a useful function called `contrast_code()` which takes
a factor column and a desired contrast scheme and returns a new contrast matrix.
The contrast scheme can be passed as either a function that generates contrast
matrices, such as `contr.sum`, or a matrix you've created yourself.

```{r}
contrast_code(mdl_data$carb, contr.sum)
```

There are two things to note here. First, the comparison labels have been retained.
This makes model intepretation easier. Second, the reference level has been
set to the first level alphabetically, rather than the last level. This is
the default behavior, but can be set to whatever you want.

```{r}
contrast_code(mdl_data$carb, contr.sum, reference_level = 4)
```

Note how the comparison labels have also changed appropriately.

The setting of reference levels only applies to schemes that utilize a reference
level. These include the following functions:

- `contr.sum`: compare each non-reference level to the grand mean
- `contr.treatment`: compare each level to the reference level (intercept = mean of reference level)
- `scaled_sum_code`: compare each level to the reference level (intercept = grand mean)

The following schemes depend on a particular order of factors. By default,
R orders levels alphabetically. If this isn't what you want, use `levels<-` to
change the ordering manually before applying contrasts coding.

- `forward_difference_code`: compare each level k to level k+1 (eg 1-2, 2-3)
- `backward_difference_code`: compare each level k to level k-1 (eg 3-2, 2-1)
- `helmert_code`: compare each level k to the mean of all levels above it (eg >1, >2)
- `reverse_helmert_code`: compare each level k to the mean of all levels below if (eg <2, <3)

Polynomial contrasts with `contr.poly` is used for trends, not group comparisons,
so the usage is a bit different. The comparison labels will still work as
expected though.

```{r}
contrast_code(mdl_data$carb, contr.poly)
```

## Utilizing `contrasts` argument in model functions

Rather than manually setting the contrasts on the data frame itself, you can
also utilize the contrasts argument in functions like `lm`, which take a named
list of contrast matrices where the names correspond to factor predictors. This
package provides a function `enlist_contrasts` to take advantage of this. Make
sure your model fitting function supports this argument, as `brm` does not use this
(see the next section for a workaround).

```{r}
my_contrasts <- 
  enlist_contrasts(mdl_data,
                   cyl ~ contr.sum + 6, # Set the reference level with + ___
                   twolevel ~ scaled_sum_code + "a",
                   gear ~ forward_difference_code,
                   carb ~ helmert_code)

my_model <- lm(mpg ~ cyl + twolevel + gear + carb, 
               data = mdl_data,  
               contrasts = my_contrasts)

summary(my_model)
```

The left hand side of each formula in `enlist_contrasts` should be the unquoted
name of a factor column, and the right hand side should be of the following:

- An unquoted function name that generates contrast matrices
- ***OR*** An unquoted variable name that's assigned to a contrast matrix
- ***OR*** A `matrix()` call, which will be captured and evaluated
- ***AND*** An optional reference level when applicable, following a `+`. String
names should be quoted, if the label happens to be numeric you can leave the number
unquoted. You can also use an unquoted variable name that contains the reference
level.

When using this method you should set contrasts for all factors. You'll receive
a warning if you don't set one or more:

```{r}
contrast_list <- enlist_contrasts(mdl_data, carb ~ helmert_code)
```

## Setting multiple contrasts at once

As previously mentioned, the `brm` function from the `brms` package (used to
fit bayesian models) does not have a `contrasts` argument. This means we need
to set the contrasts on the dataframe itself. But, this doesn't mean we need
to revert to multiple `contrasts<-` lines. We can use `set_contrasts`, which
uses the same syntax as `enlist_contrasts` but returns a new copy of the dataframe
where the factor levels are set.

```{r}
mdl_data2 <- 
  set_contrasts(mdl_data,
                cyl ~ contr.sum + 6,
                twolevel ~ scaled_sum_code + "a",
                gear ~ forward_difference_code,
                carb ~ helmert_code)

# Compare
contrasts(mdl_data$carb)
contrasts(mdl_data2$carb) %>% fractions()
```

## Note about 2 level factors

Coding factors with only 2 levels is a bit tricky conceptually because there
is only one possible comparison you can make. Thus, many of the different coding
schemes end up being mathematically equivalent.

```{r}
helmert_code(2) # Compare k to >k
scaled_sum_code(2) # Compare k to reference level
forward_difference_code(2) # Compare k to k+1
```

In many introductions to contrast coding, the motivation for using `+/-.5` is
something like this:

 - `0, 1` from `contr.treatment(2)` puts the intercept at the reference level, 
 which we don't always want
 - `-1, 1` from `-contr.sum(2)` keeps the same reference level and the intercept 
 is now the grand mean, but this tests half the difference between levels
 - `-.5, .5` from `-contr.sum(2)/2` scales the estimate so it tests the full 
 difference between levels.
 
All of this is true, but it leads to an issue when naively extending this line
of thinking beyond 2 levels. That is, one might be tempted to use the following
matrix for "sum coding a 4 level factor"

```{r}
-contr.sum(4)/2
```

Yet, this contrast matrix ***does not*** test the hypotheses that $\mu_1=\mu_4$
and $\mu_2=\mu_4$ and $\mu_3=\mu_1$ in the same way that the 2 level matrix tested
$\mu_1=\mu_2$. Rather, it tests the hypothesis that 
$1.5\mu_1 =.5\mu_2+.5\mu_3 + .5\mu_4$ etc. To actually compare, and test the 
difference between, each level and the reference level, you need to use the 
following matrix, which this package provides a helper for:

```{r}
scaled_sum_code(4) %>% fractions()
```

We can verify this by looking at the hypothesis matrices.

```{r}
# Not what we want
matrix(c(1,1,1,1,-contr.sum(4)/2), nrow = 4) %>%
  t() %>% solve() %>% fractions()

# Actually what we want
matrix(c(1,1,1,1,scaled_sum_code(4)), nrow = 4) %>%
  t() %>% solve() %>% fractions()
```

Rather than thinking about "testing half the difference," framing things in terms
of the underlying hypothesis matrix being tested is more generalizable. Assume
level 1 is `A` (and is the reference level) and level 2 is `B` in these
descriptions:

 - `0, 1` from `contr.treatment(2)` puts the intercept at the reference level
 and tests the difference $\mu_B-\mu_A$. Each comparison tests a distance from
 the reference level. We use the term *treatment coding*.
 - `-1, 1` from `-contr.sum(2)` puts the intercept at the grand mean 
 ($GM=\frac{\mu_A+\mu_B}{2}$) and tests the difference $\mu_B - GM$. Each comparison
 tests a distance from the grand mean. We use the term *sum coding*
 - `-.5, .5` from `-contr.sum(2)/2` puts the intercept at the grand mean and 
 tests the difference $\mu_B - \mu_A$. Each comparison tests a distance from
 the reference level. We use the term *scaled sum coding*.
 
It's worth noting that the term "sum coding" is not consistent across disciplines.
Other terms that are used are simple coding, deviation coding, and contrast coding.
I recommend that you choose a term from your discipline and stick to it,
but also provide the exact contrast or hypothesis matrices in an appendix,
preregistration, or supplementary material.

# Conclusion

Hopefully this vignette highlighted some useful functionality available in this
package. It does presuppose that you know a bit about what each scheme does. But,
looking at the documentation for each function provides a bit of help about
how to interpret fixed effects, eg `?scaled_sum_code`. The interpretation
of interaction terms is more complicated and beyond the scope of this vignette,
so choose your contrasts and models wisely!
